---
title: "COVID-19-Spain-Analysis"
author: "F.A. Lopez, A. Paez, Tatiane Menezes de Almeida"
date: "4/11/2020"
output:
  pdf_document: default
  html_document: default
---

This notebook reports our initial analysis of COVID-19 incidence in Spain and the climatic correlates of incidence. The data have been organized in a package for ease of access and distribution. The name of the package is `covid19env` and if necessary can be installed from the GitHub repository.

## Preliminaries

Load packages:
```{r library, collapse=TRUE, message=FALSE}
library(covid19env)
library(ggthemes)
library(gridExtra)
library(lubridate)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
#library(spatialreg)
#library(systemfit)
#library(plm)
#library(splm)
```

Load data from package `covid19env`
```{r}
data("covid19_spain")
```

Summarize the data:
```{r}
covid19_spain %>% 
  select(-geometry) %>% 
  summary()
```

The dataframe is a simple features object with information at the level of the province. The dataframe includes information about the province, including its Autonomous Community (a superior jurisdiction), an identifier, dates, COVID-19 cases and incidence. The period covered is from March 13, 2020 to April 11, 2020. In addition there are some demographic controls, and various climatic variables. Of interest are the lagged variables. The lagged variables are 8-day moving averages calculated using date-minus-12-days to date-minus-5-days, to account for the latency of the infection. More information about the dataset can be obtained by typing `?covid18_spain`.

There are 50 provinces in Spain:
```{r}
nlevels(covid19_spain$province)
```

The dataset covers 30 days:
```{r}
T <- max(covid19_spain$Date) - min(covid19_spain$Date) + 1 # To include the starting day
T
```

The order to shelter in place in Spain went into effect on March 16, 2020. March 13 is the first day that every province had at least one reported case of COVID-19.

Calculate population density and convert GDP per capita to thousands of euros:
```{r}
covid19_spain <- covid19_spain %>%
  mutate(GDPpc = GDPpc/1000,
         Density = Population/units::set_units(st_area(covid19_spain), km^2))
```

## Data exploration

Correlation analysis with Incidence:
```{r}
covid19_spain %>% 
  st_drop_geometry() %>%
  group_by(Date) %>%
  summarize(correlation_age = cor(Median_Age, Incidence),
            correlation_older = cor(Older, Incidence),
            correlation_m2f = cor(Male2Female, Incidence),
            correlation_density = cor(Density, Incidence),
            correlation_gdppc = cor(GDPpc, Incidence),
            correlation_humidity = cor(Humidity_lag11, Incidence),
            correlation_temp = cor(Mean_Temp_lag11, Incidence),
            correlation_sunshine = cor(Sunshine_Hours_lag11, Incidence)) %>%
  summary()
```

Correlation analysis with Incidence (log-transformed variables):
```{r}
covid19_spain %>% 
  st_drop_geometry() %>%
  group_by(Date) %>%
  summarize(correlation_age = cor(log(Median_Age), log(Incidence)),
            correlation_older = cor(log(Older), log(Incidence)),
            correlation_m2f = cor(log(Male2Female), log(Incidence)),
            correlation_density = cor(log(Density), log(Incidence)),
            correlation_gdppc = cor(log(GDPpc), log(Incidence)),
            correlation_humidity = cor(log(Humidity_lag11), log(Incidence)),
            correlation_temp = cor(log(Mean_Temp_lag11), log(Incidence)),
            correlation_sunshine = cor(log(Sunshine_Hours_lag11 + 0.1), log(Incidence))) %>%
  summary()
```

There are 30 days in the dataset. We can summarize the incidence by week (excluding Canarias):
```{r}
week11.plot <- covid19_spain %>% 
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~week)

week12.plot <- covid19_spain %>% 
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~week)

week13.plot <- covid19_spain %>% 
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~week)

week14.plot <- covid19_spain %>% 
  filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom") + 
  facet_wrap(~week)

grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, nrow = 2)
```

We consider some control variables: ratio of male to female in the province, median age of the population, population density:
```{r}
covid19_spain %>% 
  filter(CCAA != "Canarias", Date == "2020-03-13") %>%
  ggplot() +
  geom_sf(aes(fill = Male2Female)) +
  scale_fill_distiller(name = "Male to Female Ratio", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")
```

Median age of the population:
```{r}
covid19_spain %>% 
  filter(CCAA != "Canarias", Date == "2020-03-13") %>%
  ggplot() +
  geom_sf(aes(fill = Median_Age)) +
  scale_fill_distiller(name = "Median Age of Population", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")
```

Population density:
```{r}
covid19_spain %>% 
  filter(CCAA != "Canarias", Date == "2020-03-13") %>%
  ggplot() +
  geom_sf(aes(fill = as.numeric(Density))) +
  scale_fill_distiller(name = "Population Density", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")
```

Older people:
```{r}
covid19_spain %>% 
  filter(CCAA != "Canarias", Date == "2020-03-13") %>%
  ggplot() +
  geom_sf(aes(fill = as.numeric(Older))) +
  scale_fill_distiller(name = "Percentage of older adults", 
                       palette = "Blues", 
                       direction = 1) +
  theme_tufte() +
  theme(axis.text = element_blank(),
        legend.position = "bottom")
```

We are also interested in the climatic variables. To visualize the distribution of temperature by CCAA, we want to sort the communities by latitude, from north to south:
```{r}
# Autonomous communities
ccaa.sf <- covid19_spain %>% 
  filter(Date == "2020-03-13") %>% 
  group_by(CCAA) %>% 
  summarize(provinces = n())

# Extract coordinates of autonomous communities
ccaa.coords <- ccaa.sf %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame()

# Join Y coordinate to ccaa.sf
ccaa.sf <- ccaa.sf %>% 
  mutate(long = ccaa.coords$Y)

# Sort autonomous communities from north to south
ccaa.levels <- ccaa.sf %>% 
  arrange(desc(long)) %>% select(CCAA)

ccaa.levels <- as.character(ccaa.levels$CCAA)

# Relevel autonomous communities
covid19_spain <- covid19_spain %>%
  mutate(CCAA = factor(CCAA, levels = ccaa.levels, ordered = TRUE))
```

The following plot is the distribution of temperature by CCAA, sorted by latitude:
```{r}
# Boxplots of temperatures
covid19_spain %>% 
  st_drop_geometry() %>%
  group_by(CCAA, Date) %>%
  summarize(Mean_Temp = mean(Mean_Temp)) %>%
  ggplot(aes(x = CCAA, y = Mean_Temp)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Autonomous Community (sorted from north to south)") +
  ylab("Mean Temperature in the Community")
```

## Multivariate analysis: comparison of approaches

### Panel

1) Panel clásico

1) Panel Clásico o Dinámico

* Debe ser un modelo de efectos fijos para recoger la hetereogeneidad entre las distintas provincias (efectos)
* Debería incluir estructura dinámica ya que la serie tiene una fuerte estrutura temporal
* INCONVENIENTE: considera que la incluencia del dato del día anterior es constante (se estima un coeficiente constante)
* INCONVENIENTE: No se pueden incluir varaibles constantes en T. La hetereogeneidad entre provincias queda en el efecto fijo. No podemos por tanto incluir datos sobre estructura de la poblacion.
* INCONVENIENTE: No podemos incorporar efectos espaciales. El paquete **splm** no incluye estimación de paneles dinámicos con efectos espaciales. Tendriamos que hacerlo en matlab con los códigos de P.Elhorst.

### Spatial SUR

2) SUR espacial

* Hay un coeficente para cada variable y cada instante de tiempo. Aunque es posible considerar coeficientes constantes para los periodos temporales que consideremos.
* La hetereoeneidad espacial debemos incorporarla mediante variables explicativas. --> Estructura de la población relacionada con COVID-19.
* Permite incluir varaibles constantes en T.
* la dinámica temporar quedará recogida mediante el término independiente y la estructura de correlaciones en los residuos. EN TODO CASO, ENTIENDO QUE NUESTRO OBJETIVO NO ES EXPLICAR ESA TENEDENCIA TEMPORAL (solo modelizarla para no incurrir en errores)

## Prepare data for SUR analysis

*El modelo debe considerar efectos del 'individuo' y del 'tiempo' (para incorporar tendencia temporal)**

```{r}
# Definicion del panel para plm
GPanel <- plm::pdata.frame(covid19_spain %>% 
                             st_drop_geometry() %>%
                             select(province, 
                                    Date,
                                    Incidence, 
                                    Median_Age,
                                    Male2Female,
                                    Older,
                                    GDPpc,
                                    Density,
                                    Transit,
                                    Mean_Temp_lag8,
                                    Humidity_lag8,
                                    Sunshine_Hours_lag8,
                                    Mean_Temp_lag11,
                                    Humidity_lag11,
                                    Sunshine_Hours_lag11,
                                    Mean_Temp_lag11w,
                                    Humidity_lag11w,
                                    Sunshine_Hours_lag11w),
                           c("province","Date"))

```

### Spatial SUR model

Create connectivity matrix:
```{r}
# Extract provinces
provinces <- covid19_spain %>% 
  filter(Date == "2020-03-13") %>%
  select(province, CCAA, ID_INE)

# Spatial weights matrix: 
Wmat <- provinces %>%
  as("Spatial") %>%
  poly2nb(queen = FALSE) %>%
  nb2mat(zero.policy = TRUE)

Wmat <- (Wmat > 0) * 1

# Join the two provinces in Canarias
Wmat[which(provinces$province == "Palmas(Las)"), which(provinces$province == "Santa Cruz de Tenerife")] <- 1
Wmat[which(provinces$province == "Santa Cruz de Tenerife"), which(provinces$province == "Palmas(Las)")] <- 1

# 'Paises Catalans'
#n = 8
Wmat[which(provinces$province == "Barcelona"), which(provinces$province == "Baleares")] <- 1
Wmat[which(provinces$province == "Baleares"), which(provinces$province == "Barcelona")] <- 1
Wmat[which(provinces$province == "Baleares"), which(provinces$province == "Castellon/Castello")] <- 1 
Wmat[which(provinces$province == "Castellon/Castello"), which(provinces$province == "Baleares")] <- 1
Wmat[which(provinces$province == "Baleares"), which(provinces$province == "Tarragona")] <- 1 
Wmat[which(provinces$province == "Tarragona"), which(provinces$province == "Baleares")] <- 1
miW <- Wmat/rowSums(Wmat)

# Convert to listw
listw <- mat2listw(Wmat,style = "W")
```

Define formulas with three different lagged variables:
```{r}
formula_lag8 <- log(Incidence) ~ log(Male2Female) + 
  log(Median_Age) + 
  #log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag8) +
  log(Mean_Temp_lag8)

formula_lag11 <- log(Incidence) ~ log(Male2Female) + 
  log(Median_Age) + 
  #log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag11) +
  log(Mean_Temp_lag11) 

formula_lag11w <- log(Incidence) ~ log(Male2Female) + 
  log(Median_Age) + 
  #log(Older) +
  log(Density) +
  Transit +
  log(Humidity_lag11w) +
  log(Mean_Temp_lag11w)
```

Create the terms needed to impose restrictions to the parameters for estimation. In this case we will restrict the two demographic variables and let Density, Transit, and the climatic variables to vary across equations. The rationale is that age and ratio of male to female do not change in the short period of time examined; on the other hand, while density and the presence of transit systems are also constants over the period examined, the behavior changed as a consequence of the lockdown: we expect these variables to be significant early on in the evolution of the pandemic, and become non-significant as the lockdown reduces their importance for the transmission of the virus.
```{r}
# Recall that T is the number of days, i.e., time periods, i.e., equations
k <- 7 # Number of independent variables, including the constant
coef_rest <- 2 # Number of restrictions

# nrow is number of equations (time periods) minus 1, times the number of restrictions
# ncol is number of variables times number of equations
R2 <- matrix(0, nrow = (T - 1) * coef_rest, ncol = k * T)

for (i in 1:(T-1)){
  R2[i, 2] <- 1
  R2[i, (2 + i * k)] <- -1
  R2[(i + T - 1), 3] <- 1
  R2[(i + T - 1), (3 + i * k)] <- -1
  # Use if more restrictions are needed
  #R2[(i + T - 1) * 2, 4] <- 1
  #R2[(i + T - 1) * 2, (4 + i * k)] <- -1
}
b2 <- matrix(0, ncol = 21*coef_rest)
```

Model with a lagged 8-day moving average of climatic variables:
```{r}
sur.slm_lag8 <- spsur::spsurtime(formula = formula_lag8, 
                                 data=GPanel, 
                                 time = GPanel$Date, 
                                 type = "slm", 
                                 fit_method = "3sls", 
                                 listw=  listw,
                                 R = R2,
                                 b = b2)
summary(sur.slm_lag8)
```

Model with 11-day moving average of climatic variables:
```{r}
sur.slm_lag11 <- spsur::spsurtime(formula = formula_lag11, 
                                  data=GPanel, 
                                  time = GPanel$Date, 
                                  type = "slm", 
                                  fit_method = "3sls", 
                                  listw=  listw,
                                 R = R2,
                                 b = b2)
summary(sur.slm_lag11)
```

Model with 11-day weighted moving average of climatic variables:
```{r}
sur.slm_lag11w <- spsur::spsurtime(formula = formula_lag11w, 
                                   data=GPanel, 
                                   time = GPanel$Date, 
                                   type = "slm", 
                                   fit_method = "3sls", 
                                   listw=  listw,
                                 R = R2,
                                 b = b2)
summary(sur.slm_lag11w)
```

Compare goodness of fit:
```{r}
data.frame(R2_lag8 = sur.slm_lag8$R2, 
           R2_lag11 = sur.slm_lag11$R2,
           R2_lag11w = sur.slm_lag11w$R2) %>%
  slice(2:n()) %>%
  rownames_to_column(var = "Equation") %>%
  mutate(Date = seq(ymd("2020-03-13"), 
                    ymd("2020-04-11"), 
                    by = "days")) %>%
  pivot_longer(cols = starts_with("R"), names_to = "Model", values_to = "R2") %>%
  ggplot(aes(x = Date, y = R2, color = Model, shape = Model)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("R2_lag11w" = "blue", "R2_lag11" = "orange", "R2_lag8" = "green") ) + 
  theme_tufte()

```

## Spatial evolution of spatial residual autocorrelation

Plot the evolution of the spatial autocorrelation parameter:
```{r}
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Deltas = sur.slm_lag11$deltas,
           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
  mutate(sig = ifelse(abs(tvalue) > 1.64, "Significant at 5%", "Not Significant at 5%")) %>%
  ggplot(aes(x = Date, y = Deltas, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Significant at 5%" = "red", "Not Significant at 5%" = "blue")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

# Analisis of autocorrelated residuals

Identify all equations with significant autocorrelation parameters:
```{r}
deltas.sig <- data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Deltas = sur.slm_lag11$deltas,
           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
  mutate(sig = abs(tvalue) > 1.64)
```

Extract all residuals and calculate the spatially autocorrelated residuals:
```{r}
all_residuals <- data.frame(matrix(unlist(residuals(sur.slm_lag11)),
                                   ncol = length(residuals(sur.slm_lag11)), 
                                   byrow = FALSE))

# Spatially lagged residuals
for(i in 1:n){
  all_residuals[,i] <- deltas.sig$Deltas[i] * lag.listw(listw, all_residuals[,i])
}

# Pivot table
all_residuals <- all_residuals %>% 
  pivot_longer(everything(), 
               names_to = "Date", 
               values_to = "Spatially_Autocorrelated_Residuals")%>%
  mutate(Date = rep(seq(dmy("13-03-2020"), 
                        dmy("11-04-2020"), 
                        by = "days"), 
                    50),
         ID_INE = rep(provinces$ID_INE, 
                    each = 30))
```

Join to `covid19_spain`:
```{r}
covid19_spain <- covid19_spain %>%
  left_join(all_residuals, by = c("Date", "ID_INE"))
```

Plot residuals on March 13 (positive autocorrelation):
```{r}
covid19_spain %>% filter(Date == "2020-03-13") %>% 
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = Spatially_Autocorrelated_Residuals)) +
  scale_fill_distiller(name = "Autocorrelated Residuals", palette = "RdBu", direction = 1) +
  theme_tufte() + 
  theme(legend.position = "bottom")
```

Plot residuals on March 24 (positive autocorrelation):
```{r}
covid19_spain %>% filter(Date == "2020-03-24") %>% 
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = Spatially_Autocorrelated_Residuals)) +
  scale_fill_distiller(name = "Autocorrelated Residuals", palette = "RdBu", direction = 1) +
  theme_tufte() + 
  theme(legend.position = "bottom")
```

Plot residuals on April 11:
```{r}
covid19_spain %>% filter(Date == "2020-04-11") %>% 
  filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = Spatially_Autocorrelated_Residuals)) +
  scale_fill_distiller(name = "Autocorrelated Residuals", palette = "RdBu", direction = 1) +
  theme_tufte() + 
  theme(legend.position = "bottom")
```

Are these spatially autocorrelated residuals correlated with any other potential control variables? Check the correlations:
```{r}
covid19_spain %>% 
  st_drop_geometry() %>%
  group_by(Date) %>%
  summarize(correlation_gdppc = cor(GDPpc, Spatially_Autocorrelated_Residuals),
            correlation_older = cor(Older, Spatially_Autocorrelated_Residuals),
            correlation_sunshine = cor(Sunshine_Hours_lag11, Spatially_Autocorrelated_Residuals)) %>%
  summary()
```

These two variables are only weakly and inconsistently correlated with the residuals.



## Temporal variation of coefficients of climatic variables

Humidity:
```{r}
n = 4
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = ifelse(abs(tvalue) > 1.64, "Significant at 5%", "Not Significant at 5%")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Significant at 5%" = "red", "Not Significant at 5%" = "blue")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Temperature:
```{r}
n = 5
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = ifelse(abs(tvalue) > 1.64, "Significant at 5%", "Not Significant at 5%")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Significant at 5%" = "red", "Not Significant at 5%" = "blue")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

Boxplot of temperatures by date
```{r}
ggplot(data = covid19_spain, aes(x = Date, y = Mean_Temp_lag11, group = Date)) +
  geom_boxplot()
```

Intercept
```{r}
n = 1
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = ifelse(abs(tvalue) > 1.64, "Significant at 5%", "Not Significant at 5%")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Significant at 5%" = "red", "Not Significant at 5%" = "blue")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```